A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-25 paper.tex
Added init generalisation optimisation + more editing

\algnewcommand{\IIf}[1]{\State\algorithmicif\ #1\ \algorithmicthen}
\algnewcommand{\EndIIf}{}
\MakeRobust{\Call}
\newcommand{\True}{\texttt{true}}
\newcommand{\False}{\texttt{false}}
\newcommand{\II}{\mathcal{I}}
\newcommand{\textoverline}[1]{$\overline{\mbox{#1}}$}

@@

\begin{algorithm}
    \caption{Tree formulas for Controller and Environment respectively}
    \label{alg:treeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{gt}
        \If{$\Call{rank}{gt} == 0$}
        \State \Return{ $\lnot \Call{E}{x^{gt}}$ }
        \Else
        \State \Return{ $\lnot \Call{E}{x^{gt}} \land \bigwedge_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{treeFormula}{n})$ }
        \EndIf
        \EndFunction
        \algstore{tf1}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{tf1}
        \Function{\textoverline{treeFormula}}{gt}
        \If{$\Call{rank}{gt} == 0$}
        \State \Return{\Call{E}{$x^{gt}$}}
        \Else
        \State \Return{ $\Call{E}{x^{gt}} \lor \bigvee_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{\textoverline{treeFormula}}{n})$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$\langle s, k \rangle , T$}
        \State $cand \gets $ \Call{findCandidate}{$\langle s, k \rangle, T$} \Comment{Look for a candidate}
        \IIf{$k = n - 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$\langle s, k \rangle, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{gtAppend}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$\langle s, k \rangle, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$\langle s, k \rangle, T$}
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land$ \Call{treeFormula}{$T$} }
            \IIf{$sol = \texttt{unsat}$} 
                \Return $\emptyset$ \Comment{No candidate exists}
            \EndIIf
            \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{T} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b2}
        \Function{verify}{$\langle s, k\rangle, gt, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$\langle s, k\rangle, gt, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{\textoverline{solveAbstract}}{$\langle k', s' \rangle, \emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

\begin{algorithm}
    \caption{Amended tree formulas for Controller and Environment respectively}
    \label{alg:treeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{gt}
        \If{$\Call{rank}{gt} == 0$}
        \State \Return{ $\lnot \Call{$B^M$}{x^{gt}}$ }
        \Else
        \State \Return{ $\lnot \Call{$B^M$}{x^{gt}} \land \bigwedge_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{treeFormula}{n})$ }
        \EndIf
        \EndFunction
        \algstore{tf2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{tf2}
        \Function{\textoverline{treeFormula}}{gt}
        \If{$\Call{rank}{gt} == 0$}
        \State \Return{\Call{E}{$x^{gt}$}}
        \Else
        \State \Return{ $\Call{$B^m[rank(gt)]$}{x^{gt}} \land$ \\
            $(\Call{E}{x^{gt}} \lor \bigvee_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{$\overline{treeFormula}$}{n}))$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveUnbounded}{$T$}
            \State $B^M \gets E$
            \State $B^m[0] \gets E$
            \For{$k = 1 \dots$}
                \IIf{\Call{SAT}{$I \land B^M$}}
                    \Return \texttt{unrealisable} \Comment{Losing in the initial state}
                \EndIIf
                \IIf{$\exists i \  B^m[i] \equiv B^m[i+1]$}
                    \Return \texttt{realisable} \Comment{Reached fixed point}
                \EndIIf
                \State $B^m[k] \gets \top$
            \EndFor
        \EndFunction
        \algstore{u1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u1}
        \Require \emph{May} and \emph{must} invariants hold
        \Ensure \emph{May} and \emph{must} invariants hold
        \Ensure $\lnot s \in B^m[k]$ if there is a winning controller strategy of length $k$ starting at $s$
        \Ensure $s \in B^M$ if there is a winning environment strategy of length $k$ starting at $s$
%%%        $\exists u_{k..1} \forall c_{k..1} \  s(x_k) \land (E_k \lor (\delta(x_k, u_k, c_k) \land E(x_{k-1}) \lor ... \land E(x_1) \lor (\delta(x_1, u_1, c_1) \land E(x_0))...)$
%%%            $\forall u_{k..1} \exists c_{k..1} \  s(x_k) \land \lnot E_k \land \delta(x_k, u_k, c_k) \land \lnot E(x_{k-1}) \land ... \land \delta(x_1, u_1, c_1) \land \lnot E(x_0)$
        \Function{checkRank}{$s, k$}
            \State \Return \Call{solveAbstract}{$\langle s, k \rangle, \emptyset$}
        \EndFunction
        \algstore{u2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u2}
        \Function{solveAbstract}{$\langle s, k \rangle , T$}
        \State $cand \gets $ \Call{findCandidate}{$\langle s, k \rangle, T$} \Comment{Look for a candidate}
        \IIf{$k = n - 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$\langle s, k \rangle, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{gtAppend}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$\langle s, k \rangle, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{u3}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u3}
        \Function{findCandidate}{$\langle s, k \rangle, T$}
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land$ \Call{treeFormula}{$T$} }
            \If{$sol = \texttt{unsat}$} 
                \State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                \State \Call{learn}{$\sigma, T$}
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{T} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{u4}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u4}
        \Function{verify}{$\langle s, k\rangle, gt, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$\langle s, k\rangle, gt, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{\textoverline{solveAbstract}}{$\langle k', s' \rangle, \emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

@@

>>>>

\algnewcommand{\IIf}[1]{\State\algorithmicif\ #1\ \algorithmicthen}
\algnewcommand{\EndIIf}{}
\algnewcommand{\IfElse}[3]{\algorithmicif\ #1\ \algorithmicthen\ #2\ \algorithmicelse\ #3}
\algnewcommand{\EndIfElse}{}
\MakeRobust{\Call}
\newcommand{\True}{\texttt{true}}
\newcommand{\False}{\texttt{false}}
\newcommand{\II}{\mathcal{I}}
\newcommand{\textoverline}[1]{$\overline{\mbox{#1}}$}

@@

\begin{algorithm}
    \caption{Tree formulas for Controller and Environment respectively}
    \label{alg:treeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{gt}
        \If{$\Call{rank}{gt} = 0$}
        \State \Return{ $\lnot \Call{E}{x^{gt}}$ }
        \Else
        \State \Return{ $\lnot \Call{E}{x^{gt}} \land \bigwedge_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{treeFormula}{n})$ }
        \EndIf
        \EndFunction
        \algstore{tf1}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{tf1}
        \Function{\textoverline{treeFormula}}{gt}
        \If{$\Call{rank}{gt} = 0$}
        \State \Return{\Call{E}{$x^{gt}$}}
        \Else
        \State \Return{ $\Call{E}{x^{gt}} \lor \bigvee_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{\textoverline{treeFormula}}{n})$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$p, \langle s, k \rangle , T$}
        \State $cand \gets $ \Call{findCandidate}{$p, \langle s, k \rangle, T$} \Comment{Look for a candidate}
        \IIf{$k = n - 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$p, \langle s, k \rangle, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{gtAppend}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$p, \langle s, k \rangle, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$p, \langle s, k \rangle, T$}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$T$}}{\Call{\textoverline{treeFormula}}{$T$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land f$}
            \IIf{$sol = \texttt{unsat}$} 
                \Return $\emptyset$ \Comment{No candidate exists}
            \EndIIf
            \State \Return $\{ \langle n, m \rangle | n \in $ \Call{gtNodes}{T} $, m = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b2}
        \Function{verify}{$p, \langle s, k\rangle, gt, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$\langle s, k\rangle, gt, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{solveAbstract}{\Call{opponent}{$p$}, $\langle k', s' \rangle, \emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

\begin{algorithm}
    \caption{Amended tree formulas for Controller and Environment respectively}
    \label{alg:treeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{gt}
        \If{$\Call{rank}{gt} = 0$}
        \State \Return{ $\lnot \Call{$B^M$}{x^{gt}}$ }
        \Else
        \State \Return{ $\lnot \Call{$B^M$}{x^{gt}} \land \bigwedge_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{treeFormula}{n})$ }
        \EndIf
        \EndFunction
        \algstore{tf2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{tf2}
        \Function{\textoverline{treeFormula}}{gt}
        \If{$\Call{rank}{gt} = 0$}
        \State \Return{\Call{E}{$x^{gt}$}}
        \Else
        \State \Return{ $\Call{$B^m[rank(gt)]$}{x^{gt}} \land$ \\
            $(\Call{E}{x^{gt}} \lor \bigvee_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{$\overline{treeFormula}$}{n}))$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveUnbounded}{$T$}
            \State $B^M \gets E$
            \State $B^m[0] \gets E$
            \For{$k = 1 \dots$}
                \IIf{\Call{SAT}{$I \land B^M$}}
                    \Return \texttt{unrealisable} \Comment{Losing in the initial state}
                \EndIIf
                \IIf{$\exists i \  B^m[i] \equiv B^m[i+1]$}
                    \Return \texttt{realisable} \Comment{Reached fixed point}
                \EndIIf
                \State $B^m[k] \gets \top$
                \State \Call{checkRank}{$I, k$}
            \EndFor
        \EndFunction
        \algstore{u1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u1}
        \Require \emph{May} and \emph{must} invariants hold
        \Ensure \emph{May} and \emph{must} invariants hold
        \Ensure $\lnot s \in B^m[k]$ if there is a winning controller strategy of length $k$ starting at $s$
        \Ensure $s \in B^M$ if there is a winning environment strategy of length $k$ starting at $s$
%%%        $\exists u_{k..1} \forall c_{k..1} \  s(x_k) \land (E_k \lor (\delta(x_k, u_k, c_k) \land E(x_{k-1}) \lor ... \land E(x_1) \lor (\delta(x_1, u_1, c_1) \land E(x_0))...)$
%%%            $\forall u_{k..1} \exists c_{k..1} \  s(x_k) \land \lnot E_k \land \delta(x_k, u_k, c_k) \land \lnot E(x_{k-1}) \land ... \land \delta(x_1, u_1, c_1) \land \lnot E(x_0)$
        \Function{checkRank}{$s, k$}
            \State \Return \Call{solveAbstract}{$\texttt{env}, \langle s, k \rangle, \emptyset$}
        \EndFunction
        \algstore{u2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u2}
        \Function{solveAbstract}{$p, \langle s, k \rangle , T$}
        \State $cand \gets $ \Call{findCandidate}{$p, \langle s, k \rangle, T$} \Comment{Look for a candidate}
        \IIf{$k = n - 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$p, \langle s, k \rangle, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{gtAppend}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$p, \langle s, k \rangle, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{u3}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u3}
        \Function{findCandidate}{$p, \langle s, k \rangle, T$}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$T$}}{\Call{\textoverline{treeFormula}}{$T$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land f$}
            \If{$sol = \texttt{unsat}$} 
                \State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                \State \Call{learn}{$\sigma, T$}
                \State \IfElse{$p = \texttt{cont}$}{\Call{learn}{$\sigma, T$}}{\Call{\textoverline{learn}}{$\sigma, T$}} \EndIfElse
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{T} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{u4}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u4}
        \Function{verify}{$p, \langle s, k\rangle, gt, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$\langle s, k\rangle, gt, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{solveAbstract}{\Call{opponent}{$p$}, $\langle k', s' \rangle, \emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

In this section we present some optimisations to the unbounded synthesis algorithm.

@@

If more states are removed from $B^m$ on each iteration of \textsc{checkRank}
the algorithm will converge and terminate sooner. Additionally, any states
removed from $B^M$ will reduce the number of states to be considered by the
controller in future iterations.

As it is written, the algorithm only considers an overapproximations of states
reachable from $I$ for learning. Assuming that the algorithm does not
terminate, then the addition of an extra step into the game on each iteration
has the potential to greatly increase the reachable states. Considering some of
these states earlier than they become reachable may lead to earlier
termination.

\begin{algorithm}
    \begin{algorithmic}
        \Function{checkRank}{$s, k$}
            \State $r \gets $ \Call{solveAbstract}{$\texttt{env}, \langle s, k \rangle, \emptyset$}
            \IIf{$r \neq \emptyset$} \Return $r$ \EndIIf
            \State $s' \gets s$
            \For{$x \in s$}
                \State $r \gets$ \Call{solveAbstract}{$\texttt{env}, \langle s' \setminus \{x\}, k \rangle, \emptyset$}
                \IIf{$r = \emptyset$} $s' \gets s' \setminus \{x\}$ \EndIIf
            \EndFor
            \State \Return $\emptyset$
        \EndFunction
    \end{algorithmic}
    \caption{Generalise $I$ optimisation}
    \label{alg:opt1}
\end{algorithm}

The optimisation that allows this is relatively simple and is inspired by a
common greedy heuristic for minimising $\texttt{unsat}$ cores. $I$ is a value
assignment to each variable in $X$. If the environment does not win for
$\langle I, k \rangle$ then we attempt to solve for a generalised version of
$I$ by removing one assignment at a time. If the controller can win from the
larger set of states then we continue generalising without it. In this way we
learn more states by increasing the reachable set. In our benchmarks we have
observed that this optimisation is beneficial on the first few iterations of
\textsc{checkRank}.
